---
title: "HW5_r"
output: pdf_document
date: "2026-01-23"
---

Jona Nakai
DS 4420
Homework #5

## Problem 4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(ggplot2)
```

```{r data}
spotify <- read.csv('ds4420_spotify.csv')
head(spotify)

explicit <- spotify$explicit
n <- length(explicit)
y <- sum(explicit)
alpha <- 2
beta <- 18
```

```{r distribution}
prior <- function(theta, alpha, beta) {
  log_num <- (alpha - 1) * log(theta) + (beta - 1) * log(1 - theta)
  log_den <- lbeta(alpha, beta)
  exp(log_num - log_den)
}

posterior <- function(theta, alpha, beta, y, n) {
  alpha_star <- alpha + y
  beta_star <- beta + n - y
  log_num <- (alpha_star - 1) * log(theta) + (beta_star - 1) * log(1 - theta)
  log_den <- lbeta(alpha_star, beta_star)
  exp(log_num - log_den)
}
```

```{r plot_1}
sim_theta <- seq(0.001, 0.999, length.out = 1000)
prior_dist <- prior(sim_theta, alpha, beta)
post_dist <- posterior(sim_theta, alpha, beta, y, n)

plot_df <- data.frame(theta = sim_theta, Prior = prior_dist, Posterior = post_dist)
ggplot(plot_df, aes(x = theta)) +
  geom_line(aes(y = Prior, color = 'Prior Distribution')) +
  geom_line(aes(y = Posterior, color = 'Posterior Distribution')) +
  labs(x = expression(theta), y = expression(p(theta))) +
  scale_color_manual(name = "Legend", values = c(
    "Prior Distribution" = "green", 
    "Posterior Distribution" = "orange"
  ))

```

```{r plot_2}
prior_mean <- alpha / (alpha + beta)
MLE <- y / n
posterior_mean <- (alpha + y) / (alpha + beta + n)

prior_mean
MLE
posterior_mean

ggplot(plot_df, aes(x = theta)) +
  geom_line(aes(y = Prior, color = "Prior Distribution")) +
  geom_line(aes(y = Posterior, color = "Posterior Distribution")) +
  geom_vline(aes(xintercept = prior_mean, color = "Prior Mean")) +
  geom_vline(aes(xintercept = MLE, color = "MLE")) +
  geom_vline(aes(xintercept = posterior_mean, color = "Posterior Mean")) +
  labs(x = expression(theta), y = expression(p(theta))) +
  scale_color_manual(name = "Legend", values = c(
    "Prior Distribution" = "green", 
    "Posterior Distribution" = "orange",
    "Prior Mean" = "red",
    "MLE" = "blue",
    "Posterior Mean" = "purple"
  ))
```

The posterior mean being in between the prior and MLE makes sense as it is a weighted average of the prior and MLE. As n, 
the number of observations in our data set, grows, our posterior will converge to our MLE. The prior mean is just
our initial belief before using the data, while the MLE represents the data. We would use our posterior mean as our 
estimate for theta. So the probability of a song being explicit would be our posterior mean. To validate the performance
of our model, we can train test split the data, use the train data to create a posterior, and compare the posterior mean
to the percentage of explicit music in our test data set.


## Problem 5 Part

```{r part_a}
track_pop <- spotify$track_pop
n <- length(track_pop)
y_bar <- mean(track_pop)
sigma <- 20
eta <- 50
tao <- 2
mu_star <- (((n * y_bar) / sigma^2) + (eta / tao^2)) / ((n / sigma^2) + (1 / tao^2))
sigma_star <- sqrt(((n / sigma^2) + (1 / tao^2))^-1)

prior <- function(mu, eta, tao) {
  (1 / (tao * sqrt(2 * pi))) * exp(-(mu - eta)^2 / (2 * tao^2))
}

posterior <- function(mu, mu_star, sigma_star) {
  prior(mu, mu_star, sigma_star)
}

prior_mean <- eta
MLE <- y_bar
posterior_mean <- mu_star

prior_mean
MLE
posterior_mean

sim_mu <- seq(40, 80, length.out = 1000)
prior_dist <- prior(sim_mu, eta, tao)
post_dist <- posterior(sim_mu, mu_star, sigma_star)

plot_df <- data.frame(mu = sim_mu, Prior = prior_dist, Posterior = post_dist)
ggplot(plot_df, aes(x = mu)) +
  geom_line(aes(y = Prior, color = "Prior Distribution")) +
  geom_line(aes(y = Posterior, color = "Posterior Distribution")) +
  geom_vline(aes(xintercept = prior_mean, color = "Prior Mean")) +
  geom_vline(aes(xintercept = MLE, color = "MLE")) +
  geom_vline(aes(xintercept = posterior_mean, color = "Posterior Mean")) +
  labs(x = expression(mu), y = expression(p(mu))) +
  scale_color_manual(name = "Legend", values = c(
    "Prior Distribution" = "green",
    "Posterior Distribution" = "orange",
    "Prior Mean" = "red",
    "MLE" = "blue",
    "Posterior Mean" = "purple"
  ))
```

The posterior mean is close to the MLE as the large dataset weighs the posterior to be closer to the MLE. The prior mean
is our initial belief before data, the MLE reflects our data, and the posterior utilizes both our initial belief and 
our data.

```{r part_b_1}
eta <- 75
tao <- 2
mu_star <- (((n * y_bar) / sigma^2) + (eta / tao^2)) / ((n / sigma^2) + (1 / tao^2))
sigma_star <- sqrt(((n / sigma^2) + (1 / tao^2))^-1)

prior <- function(mu, eta, tao) {
  (1 / (tao * sqrt(2 * pi))) * exp(-(mu - eta)^2 / (2 * tao^2))
}

posterior <- function(mu, mu_star, sigma_star) {
  prior(mu, mu_star, sigma_star)
}

prior_mean <- eta
MLE <- y_bar
posterior_mean <- mu_star

prior_mean
MLE
posterior_mean

sim_mu <- seq(40, 80, length.out = 1000)
prior_dist <- prior(sim_mu, eta, tao)
post_dist <- posterior(sim_mu, mu_star, sigma_star)

plot_df <- data.frame(mu = sim_mu, Prior = prior_dist, Posterior = post_dist)
ggplot(plot_df, aes(x = mu)) +
  geom_line(aes(y = Prior, color = "Prior Distribution")) +
  geom_line(aes(y = Posterior, color = "Posterior Distribution")) +
  geom_vline(aes(xintercept = prior_mean, color = "Prior Mean")) +
  geom_vline(aes(xintercept = MLE, color = "MLE")) +
  geom_vline(aes(xintercept = posterior_mean, color = "Posterior Mean")) +
  labs(x = expression(mu), y = expression(p(mu))) +
  scale_color_manual(name = "Legend", values = c(
    "Prior Distribution" = "green",
    "Posterior Distribution" = "orange",
    "Prior Mean" = "red",
    "MLE" = "blue",
    "Posterior Mean" = "purple"
  ))
```

The MLE is now on the left side because our eta value is larger than our MLE. The posterior mean still leans close to
our MLE. 

```{r part_b_2}
eta <- 50
tao <- 5
mu_star <- (((n * y_bar) / sigma^2) + (eta / tao^2)) / ((n / sigma^2) + (1 / tao^2))
sigma_star <- sqrt(((n / sigma^2) + (1 / tao^2))^-1)

prior <- function(mu, eta, tao) {
  (1 / (tao * sqrt(2 * pi))) * exp(-(mu - eta)^2 / (2 * tao^2))
}

posterior <- function(mu, mu_star, sigma_star) {
  prior(mu, mu_star, sigma_star)
}

prior_mean <- eta
MLE <- y_bar
posterior_mean <- mu_star

prior_mean
MLE
posterior_mean

sim_mu <- seq(40, 80, length.out = 1000)
prior_dist <- prior(sim_mu, eta, tao)
post_dist <- posterior(sim_mu, mu_star, sigma_star)

plot_df <- data.frame(mu = sim_mu, Prior = prior_dist, Posterior = post_dist)
ggplot(plot_df, aes(x = mu)) +
  geom_line(aes(y = Prior, color = "Prior Distribution")) +
  geom_line(aes(y = Posterior, color = "Posterior Distribution")) +
  geom_vline(aes(xintercept = prior_mean, color = "Prior Mean")) +
  geom_vline(aes(xintercept = MLE, color = "MLE")) +
  geom_vline(aes(xintercept = posterior_mean, color = "Posterior Mean")) +
  labs(x = expression(mu), y = expression(p(mu))) +
  scale_color_manual(name = "Legend", values = c(
    "Prior Distribution" = "green",
    "Posterior Distribution" = "orange",
    "Prior Mean" = "red",
    "MLE" = "blue",
    "Posterior Mean" = "purple"
  ))
```

Compared to the other two, our posterior mean is closest to our MLE. This is due to a wider prior distribution, allowing
for the posterior distribution to be pulled towards our MLE much more.

### Part_c

The prior distribution with eta = 75 and tao = 2 resulted in the more significantly differnet posterior. This was due 
to the prior mean being shifted so far from the MLE. This exercised helped me understand how shifting parameters results
in different posteriors. Not necessarily better or worse, but different. By shifting the parameters, different posteriors
result. Some more or less influenced by the prior. Some lean the posterior in different directions by shifting the prior
mean. In practice, I would use a train test split to choose different hyperparameters and evaluate them to help me choose
the best possible hyperparameters.